State of the Practice for Medical Imaging Software Based on Open Source Repositories

### 2024-06-13 StateOfPractice/Papers/SOP_MI_OSR/SOP_MI_OSR.tex
done up to section 2.3. Basically down 2 pages.

\abstract{
We present the state of the practice for Medical Imaging (MI) software based on
data available in open source repositories. We selected 29 medical imaging
projects from 48 candidates and assessed 9 software qualities (installability,
correctness/ verifiability, reliability, robustness, usability, maintainability,
reusability, understandability, and visibility/transparency) by answering 108
questions for each software project. Based on the quantitative data, we ranked
the MI software with the Analytic Hierarchy Process (AHP). The top four software
products are \textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji}, and
\textit{OHIF Viewer}.  Our ranking is consistent with the community's
ranking, with four of our top five projects also appearing in the top five of a
list ordered by stars-per-year.  We observed 88\% of the documentation artifacts
recommended by research software development guidelines.  However, the current
state of the practice deviates from the existing guidelines because of the
rarity of some recommended artifacts (like test plans, requirements
specification, code of conduct, code style guidelines, product roadmaps, and
Application Program Interface (API) documentation).
}

@@

We aim to study the state of software development practice for Medical Imaging
(MI) software using data available in open source repositories.  MI tools use
images of the interior of the body (from sources such as Magnetic Resonance
Imaging (MRI), Computed Tomography (CT), Positron Emission Tomography (PET) and
Ultrasound) to provide information for diagnostic, analytic, and medical
applications. Through MI medical practitioners and
researchers can noninvasively gain insights into the human body, including
information on injuries and illnesses. Given the importance of MI software and
the high number of competing software projects, we wish to understand the merits
and drawbacks of the current development processes, tools, and methodologies. We
aim to assess through a software engineering lens the quality of the existing
software with the hope of highlighting standout examples, and providing
guidelines and recommendations for future development.

@@

Not only do we wish to gain insight into the state of the practice for MI
software, we also wish to understand the impact of the often cited gap, or
chasm, between software engineering and research software \cite{Storer2017}. Although scientists spend a substantial proportion of their working
hours on software development \cite{Hannay2009, Prabhu2011}, many developers
learn software engineering skills by themselves or from their peers, instead of
from proper training \cite{Hannay2009}. In the past many scientists have shown
ignorance and indifference to standard software engineering concepts
\cite{Hannay2009}. For instance, according to a survey by Prabhu in 2011
\cite{Prabhu2011}, more than half of their 114 subjects did not use a proper
debugger when coding.

To gain insight, we devised 10 research questions, which can be applied to MI,
as well as to other domains, of research software \cite{SmithEtAl2021,
SmithAndMichalski2022}.  We designed the questions to learn about the
community's interest in, and experience with, software artifacts, tools,
principles, processes, methodologies, and qualities.  When we mention artifacts
we mean the documents, scripts and code that constitutes a software development
project. Example artifacts include requirements, specifications, user manuals,
unit tests, system tests, usability tests, build scripts, API (Application
Programming Interface) documentation, READMEs, license documents, process
documents, and code.  Once we have learned what MI developers do, we then put
this information in context by contrasting MI software against the trends shown
by developers in other research software communities.

We based the structure of the paper on the research questions, so for each
research question below we point to the section that contains our answer.  We
start with identifying the relevant examples of MI software for the assessment
exercise:

\begin{enumerate}
	\item[RQ\refstepcounter{rqnum}\therqnum \label{RQ_WhatProjects}:] What MI
	software projects exist, with the constraint that the source code must be
	available for all identified projects? (Section~\ref{ch_results})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_HighestQuality}:] Which
	of the projects identified in \rqref{RQ_WhatProjects} follow current best
	practices, based on evidence found by experimenting with the software and
	searching the artifacts available in each project's repository?
	(Section~\ref{ch_results})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareHQ2Popular}:] How
	similar is the list of top projects identified in \rqref{RQ_HighestQuality}
	to the most popular projects, as viewed by the scientific community?
	(Section~\ref{Sec_VsCommunityRanking})
    \item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareArtifacts}:] How
	do MI projects compare to research software in general with respect to the
	artifacts present in their repositories?
	(Section~\ref{Sec_CompareArtifacts})
\end{enumerate}

@@

To make the project feasible, we only cover MI visualization software.  As a
consequence we are excluding many other categories of MI software, including
Segmentation, Registration, Visualization, Enhancement, Quantification,
Simulation, plus MI archiving and telemedicine systems (Compression, Storage,
and Communication).  We also exclude Statistical Analysis and Image-based
Physiological Modelling and Feature Extraction, Classification, and
Interpretation. Software that provides MI support functions is also out of
scope; therefore, we have not assessed the toolkit libraries VTK and ITK.
Finally, Picture Archiving and Communication System (PACS), which helps users to
economically store and conveniently access images, are considered out of scope.

@@

We designed a general methodology to assess the state of the practice for
research software \cite{SmithEtAl2021, SmithAndMichalski2022}.  Our methodology
has been applied to MI software and Lattice Boltzmann Solvers
\cite{SmithEtAl2024}.  This methodology builds off prior work to assess the
state of the practice for such domains as Geographic Information Systems
\cite{smith2018state}, Mesh Generators \cite{smith2016state}, Seismology
software \cite{Smith2018Seismology}, and Statistical software for psychology
\cite{smith2018statistical}.  In keeping with the previous methodology, we have
maintained the constraint that the work load for measuring a given domain should
be feasible for a team as small as one person, and for a short time, ideally
around a person month of effort. We consider a person month as $20$ working days
($4$ weeks in a month, with $5$ days of work per week) at $8$ person hours per
day, or $20 \times 8 = 160$ person hours.

With our methodology, we first choose a research software domain (in the current
case MI) and identify a list of about 30 software packages. (For measuring MI we
used 29 software packages.)  The selected packages are vetted by a domain
expert. We approximately measure the qualities of each package by filling in a
grading template. The grader looks at the contents of the open source repository
and repository based metrics, such as the number of files, number of lines of
code, percentage of issues that are closed, etc.  With the quantitative data in
the grading template, we rank the software with the Analytic Hierarchy Process
(AHP). Further details on interaction with the domain expert, software
qualities, grading the software and AHP are found below and in
\cite{SmithEtAl2024_MI_SOP}.

\subsubsection{Interaction With Domain Expert} \label{sec_vet_software_list}

The Domain Experts vet the proposed list of software packages and the AHP
ranking.  Doing these tasks with non-experts runs the risk of trusting
inaccurate resources and overlooking the tacit knowledge of an expert. For the
current assessment, our Domain Expert (and paper co-author) is Dr.\ Michael
Noseworthy, Professor of Electrical and Computer Engineering at McMaster
University, Co-Director of the McMaster School of Biomedical Engineering, and
Director of Medical Imaging Physics and Engineering at St.\ Joseph's Healthcare,
Hamilton, Ontario, Canada.

In advance of the first meeting with the Domain Expert, they are asked to
create a list of top software packages in the domain.  This is done to help
the expert get in the right mind set in advance of the meeting.  Moreover,
by doing the exercise in advance, we avoid the potential pitfall of the expert
approving the discovered list of software without giving it adequate thought.

@@

\subsubsection{Grading Software} \label{sec_grading_software}

We grade the selected software using an existing measurement template
\cite{SmithEtAl2021}.  The template provides measures of the qualities listed in
Section~\ref{sec_software_quality}.  For each software package, we fill in the
template questions. To stay within the target of 160 person hours to measure the
domain, we allocated between one and four hours for each package. Project
developers can be contacted for help regarding installation, if necessary, but
we impose a cap of about two hours on the installation process, to keep the
overall measurement time feasible.  Figure~\ref{fg_grading_template_example}
shows an excerpt of the spreadsheet.  The spreadsheet includes a column for each
measured software package. \cite{Dong2021-Data} provides the full set of
measurement data.

@@

The full template consists of 108 questions categorized under 9 qualities.  We
designed the questions to be unambiguous, quantifiable, and measurable with
limited time and domain knowledge. We group the measures under headings for each
quality, and one for summary information. The summary information (shown in
Figure~\ref{fg_grading_template_example}) is the first section of the template.
This section summarizes general information, such as the software name, purpose,
platform, programming language, publications about the software, the first
release and the most recent change date, website, source code repository of the
product, number of developers, etc.  We only measure open source software that
is ready for public use by other researchers (following the definition of the
public category from \cite{GewaltigAndCannon2012}). Information in the summary
section sets the context for the project, but it does not directly affect the
grading scores.

For measuring each quality, we ask several questions and the typical answers are
among the collection of ``yes'', ``no'', ``n/a'', ``unclear'', a number, a
string, a date, a set of strings, etc. The grader assigns each quality an
overall score, between 1 and 10, based on all the previous questions.  Several
of the qualities use the word ``surface''.  This is to highlight that, for these
qualities in particular, the best that we can do is a shallow measure.  For
instance, we are not currently doing any experiments to measure usability.
Instead, we are looking for an indication that the developers considered
usability.  We do this by looking for cues in the documentation, like a getting
started manual, a user manual and a statement of expected user characteristics.
We use two freeware tools to collect repository related data:
\href{https://github.com/tomgi/git_stats}{GitStats} and
\href{https://github.com/boyter/scc}{Sloc Cloc and Code (scc)}.  Details on how
we measure each quality are provided in Smith et al.\ (2024)
\cite{SmithEtAl2024_MI_SOP}.

@@

Saaty developed AHP in the 1970s, and people have widely used it since to make
and analyze multiple criteria decisions \cite{VaidyaEtAl2006}. AHP organizes
multiple criteria in a hierarchical structure and uses pairwise comparisons
between alternatives to calculate relative ratios \cite{Saaty1990}. AHP works
with sets of $n$ \textit{options} and $m$ \textit{criteria}.  In our project
$n=29$ and $m=9$ since there are 29 options (software products) and 9 criteria
(qualities). With AHP the sum of the grades (scores) for all products for a
given quality will be 1.0.  We rank the software for each of the qualities, and
then we combine the quality rankings into an overall ranking based on the
relative priorities between qualities.

@@

In this section we review the selected software packages.  We used a two-step
process for selecting the software packages: i) identify software candidates in
the chosen domain; and, ii) filter the list to remove less relevant members
\cite{SmithEtAl2021}.

We initially identified 48 MI candidate software projects \cite{Dong2021} from
the literature \cite{Bjorn2017, Bruhschwein2019, Haak2015}, on-line articles
\cite{Emms2019, Hasan2020, Mu2019}, and forum discussions \cite{Samala2014}.  To
reduce the length of the list to a manageable number (29 in this case), we
filtered the original list as follows:

@@

\item We removed the packages with no source code available, such as
\textit{MicroDicom}, \textit{Aliza}, and \textit{jivex}.

\item We focused on the MI software that provides visualization functions, as
described in Section~\ref{sec_scope}. Furthermore, we removed seven packages
that were toolkits or libraries, such as \textit{VTK}, \textit{ITK}, and
\textit{dcm4che}.  We removed another three that were for PACS.

\item We removed \textit{Open Dicom Viewer}, since it has not received any
updates in a long time (since 2011).

@@

The Domain Expert provided a list of his top 12 software packages.  We compared
his list to our list of 29.  We found 6 packages were on both lists: \textit{3D
Slicer}, \textit{Horos}, \textit{ImageJ}, \textit{Fiji}, \textit{MRIcron} (we
actually use the update version \textit{MRIcroGL}) and \textit{Mango} (we
actually use the web version \textit{Papaya}).  Six software packages
(\textit{AFNI}, \textit{FSL}, \textit{Freesurfer}, \textit{Tarquin},
\textit{Diffusion Toolkit}, and \textit{MRItrix}) were on the Domain Expert
list, but not on our filtered list.  When we examined those packages,
we found they were out of scope, since their primary function was not
visualization.  The Domain Expert agreed with our final choice of 29 packages.

Table~\ref{tab_final_list} shows the 29 software packages that we measured,
along with summary data collected in the year 2020. We arrange the items in
descending order of LOC. We found the initial release dates (Rlsd) for most
projects and marked the two unknown dates with ``?''. The date of the last
update is the date of the latest update, at the time of measurement. We found
funding information (Fnd) for only eight projects.  For the Number Of
Contributors (NOC) we considered anyone who made at least one accepted commit as
a contributor. The NOC is not usually the same as the number of long-term
project members, since many projects received change requests and code from the
community.  With respect to the OS, 25 packages work on all three OSs: Windows
(W), macOS (M), and Linux (L). Although the usual approach to cross-platform
compatibility was to work natively on multiple OSes, five projects achieved
platform-independence via web applications. The full measurement data for all
packages is available on \href{https://data.mendeley.com/datasets/k3pcdvdzj2/1}
{Mendeley Data}.

@@

The programming languages used in order of decreasing popularity are \CC,
JavaScript, Java, C, Python, Pascal, Matlab.  The most popular language is \CC,
with almost 40\% of projects (11 of 29).  The two least popular choices are
Pascal and Matlab, with around 3\% of projects each (1 of 29).

@@

Figure \ref{fg_installability_scores} lists the installability scores.  We found
installation instructions for 16 projects. Among the ones without instructions,
\textit{BioImage Suite Web} and \textit{Slice:Drop} do not need installation,
since they are web applications. Installing 10 of the projects required extra
dependencies. Five of them are web applications (as shown in
Table~\ref{tab_final_list}) and depend on a browser; \textit{dwv}, \textit{OHIF
Viewer}, and \textit{GATE} needs extra dependencies to build; \textit{ImageJ}
and	\textit{Fiji} need an unzip tool; \textit{MatrixUser} is based on Matlab;
\textit{DICOM Viewer} needs to work on a Nextcloud platform.

@@

\textit{3D Slicer} has the highest score because it had easy to follow
installation instructions, and an automated, fast, frustration-free installation
process. The installer added all dependencies automatically and no errors
occurred during the installation and uninstallation steps. Many other software
packages also had installation instructions and automated installers.  We had no
trouble installing the following packages: \textit{INVESALIUS 3},
\textit{Gwyddion}, \textit{XMedCon}, and \textit{MicroView}. We determined their
scores based on the understandability of the instructions, installation steps,
and user experience. Since \textit{BioImage Suite Web} and \textit{Slice:Drop}
needed no installation, we gave them high scores. \textit{BioImage Suite Web}
also provided an option to download cache for offline usage, which was easy to
apply.

\textit{GATE}, \textit{dwv}, and \textit{DICOM Viewer} showed severe
installation problems. We were not able to install them, even after a reasonable
amount of time (2 hours).  For \textit{dwv} and \textit{GATE} we failed to build
from the source code, but we were able to proceed with measuring other qualities
using a deployed on-line version for \textit{dwv}, and a VM version for
\textit{GATE}. For \textit{DICOM Viewer} we could not install the NextCloud
dependency, and we did not have another option for running the software.
Therefore, for \textit{DICOM Viewer} we could not measure reliability or
robustness.  The other seven qualities could be measured, since they do not
require installation.

\textit{MatrixUser} has a lower score because it depends on Matlab. We assessed
the score from the point of view of a user that would have to install Matlab and
acquire a license.  Of course, for users that already work within Matlab, the
installability score should be higher.

@@

Figure~\ref{fg_correctness_verifiability_scores} shows the scores of correctness
and verifiability. Generally speaking, the packages with higher scores adopted
more techniques to improve correctness, and had better documentation for us to
verify against.  For instance, we looked for evidence of unit testing, since it
benefits most parts of the software's life cycle, such as designing, coding,
debugging, and optimization \cite{Hamill2004}.  We only found evidence of unit
testing for about half of the projects. We identified five projects using CI/CD
tools: \textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji}, \textit{dwv}, and
\textit{OHIF Viewer}.

@@

Even for some projects with well-organized documentation, requirements
specifications and theory manuals were still missing.  We could not identify
theory manuals for all projects, and we did not find requirements specifications
for most projects. The only requirements-related document we found was a road
map of \textit{3D Slicer}, which contained design requirements for upcoming
changes.

@@

Figure~\ref{fg_reliability_scores} shows the AHP results.  As shown in
Section~\ref{sec_result_installability}, most of the software products did not
``break'' during installation, or did not need installation; \textit{dwv} and
\textit{GATE} broke in the building stage, and the processes were not
recoverable; we could not install the dependency for \textit{DICOM Viewer}. Of
the seven software packages with a getting started tutorial and operation steps
in the tutorial, most showed no error when we followed the steps. However,
\textit{GATE} could not open macro files and became unresponsive several times,
without any descriptive error message. We found that \textit{Drishti} crashed
when loading damaged image files, without showing any descriptive error message.
We did not find any problems with the on-line version of \textit{dwv}.

@@

Figure \ref{fg_robustness_scores} presents the scores for surface robustness.
The packages with higher scores elegantly handled unexpected/unanticipated
inputs, typically showing a clear error message. We may have underestimated the
score of \textit{OHIF Viewer}, since we needed further customization to load
data.

>>>>

\abstract{
We present the state of the practice for Medical Imaging (MI) software based on
data available in open source repositories. We selected 29 
projects from 48 candidates and assessed 9 software qualities (installability,
correctness/ verifiability, reliability, robustness, usability, maintainability,
reusability, understandability, and visibility/transparency) by answering 108
questions for each. Using the Analytic Hierarchy Process (AHP) on the
quantitive data, we ranked the MI software.  The top five are \textit{3D
Slicer}, \textit{ImageJ}, \textit{Fiji}, \textit{OHIF Viewer}, and
\textit{ParaView}.  This is consistent with the community's, with four of
these also appearing in the top five using GitHub metrics (stars-per-year).  
In general, the quality and quantity of documentation present in the
project correlates quite well with its popularity.
%We observed 88\% of the documentation artifacts
%recommended by research software development guidelines.  However, the current
%state of the practice deviates from the existing guidelines because of the
%rarity of some recommended artifacts (like test plans, requirements
%specification, code of conduct, code style guidelines, product roadmaps, and
%Application Program Interface (API) documentation).
}

@@

We study the state of software development practice for Medical Imaging
(MI) software using data available in open source repositories.  MI tools use
images of the interior of the body (from sources such as Magnetic Resonance
Imaging (MRI), Computed Tomography (CT), Positron Emission Tomography (PET) and
Ultrasound) to provide information for diagnostic, analytic, and medical
applications. Given its importance, we want to understand the merits and
drawbacks of the current development processes, tools, and methodologies. We
use a software engineering lens to assess the quality of existing MI software
and provide recommendations for future development.

@@

As well as state of the practice (SOP) for MI software, we would like to
understand the impact of the often cited gap (chasm!) between recommended
software engineering practices and that used for most research 
software~\cite{Storer2017}. Although scientists spend
a substantial proportion of their working times on software development
\cite{Hannay2009, Prabhu2011}, few are formally trained~\cite{Hannay2009}.

%We have a standard set of questions designed to assess the qualities of any
%research software project \cite{SmithEtAl2021, SmithAndMichalski2022}. We
%particularly focus on \emph{software artifactcs}, i.e.
%the documents, scripts and code that constitutes a software development
%project. Example artifacts include requirements, specifications, user manuals,
%unit tests, system tests, usability tests, build scripts, API (Application
%Programming Interface) documentation, READMEs, license documents, process
%documents, and code.  %Once we have learned what MI developers do, we then put
%% this information in context by contrasting MI software against the trends shown
%% by developers in other research software communities.

Our investigation is based on the following four research questions:
\begin{enumerate}
\item[RQ\refstepcounter{rqnum}\therqnum \label{RQ_WhatProjects}:] What MI
\textbf{open source} software projects exist? (Section~\ref{ch_results})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_HighestQuality}:] Establish
quantitative measurements of those project's development practices. From
these measures, which of those projects follow best practices?
(Section~\ref{ch_results})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareHQ2Popular}:] How
similar are the top projects identified in \rqref{RQ_HighestQuality}
to the most popular projects as viewed by the community?
(Section~\ref{Sec_VsCommunityRanking})
\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareArtifacts}:] How
do MI projects compare to general research software with respect to the
artifacts present in their repositories?
(Section~\ref{Sec_CompareArtifacts})
\end{enumerate}

@@

We only cover MI visualization software.  We exclude
other categories of MI software, including
Segmentation, Registration, Visualization, Enhancement, Quantification,
Simulation, plus MI archiving and telemedicine systems (Compression, Storage,
and Communication).  We also exclude Statistical Analysis and Image-based
Physiological Modelling and Feature Extraction, Classification, and
Interpretation. Software that provides MI support functions is also out of
scope; therefore, we have not assessed the toolkit libraries VTK and ITK.
Finally, Picture Archiving and Communication System (PACS), which helps users to
economically store and conveniently access images, are considered out of scope.

@@

We have a standard set of questions designed to assess the qualities of any
research software project~\cite{SmithEtAl2021, SmithAndMichalski2022}.
This has been applied to MI software and Lattice Boltzmann Solvers
\cite{SmithEtAl2024}.  This builds off prior work to assess the
state of the practice for such domains as Geographic Information Systems
\cite{smith2018state}, Mesh Generators \cite{smith2016state}, Seismology
software \cite{Smith2018Seismology}, and Statistical software for psychology
\cite{smith2018statistical}.  We maintain the previous constraint that the
work load for measuring a given domain should take around one person-month's
worth of effort ($20$ working days at $8$ person-hours per day).

We identify a list of potential packages (through online searches) which is
then filtered and vetted by a domain expert. We aim for roughly $30$
packages. For each remaining package, we measure its qualities by filling in a
grading template~\cite{SmithEtAl2021}.  This data is used to rank the projects
with AHP.  Further details on interaction with the domain expert, software
qualities, grading the software and AHP are summarizes below and in longer form
in~\cite{SmithEtAl2024_MI_SOP}.

\subsubsection{Domain Expert} \label{sec_vet_software_list}

The Domain Expert vets the proposed list. For too many domains, there are
online resources that are wildly inaccurate, which only an expert can correct.
(Details of our domain expert removed for double-blind).
%The Domain Experts vet the proposed list of software packages and the AHP
%ranking.  Doing these tasks with non-experts runs the risk of trusting
%inaccurate resources and overlooking the tacit knowledge of an expert. For the
%current assessment, our Domain Expert (and paper co-author) is Dr.\ Michael
%Noseworthy, Professor of Electrical and Computer Engineering at McMaster
%University, Co-Director of the McMaster School of Biomedical Engineering, and
%Director of Medical Imaging Physics and Engineering at St.\ Joseph's Healthcare,
%Hamilton, Ontario, Canada.

In advance of the first meeting with the Domain Expert, they are asked to
independently create a list of top software packages in the domain.  This helps
get the expert's knowledge refreshed in advance of the meeting.

@@

\subsubsection{Grading} \label{sec_grading_software}

We use an existing template~\cite{SmithEtAl2021} that is designed to measure
the aforementioned qualities. To stay within our given measurement time frame,
each package gets up to five hours of time.  Project developers can be
contacted for help regarding installation, if necessary, but we impose a cap of
about two hours on the installation process.
Figure~\ref{fg_grading_template_example} shows an excerpt of the measurement
spreadsheet.  It includes a column for each measured software package.
% \cite{Dong2021-Data} provides the full set of measurement data.
(The full data is available on Mendeley; link will be provided after
refereeing.)

@@

The full template consists of 108 questions over 9 qualities.  These
questions are designed to be unambiguous, quantifiable, and measurable
with constrained time and domain knowledge.
%We
%designed the questions to be unambiguous, quantifiable, and measurable with
%limited time and domain knowledge. We group the measures under headings for each
%quality, and one for summary information. The summary information (shown in
%Figure~\ref{fg_grading_template_example}) is the first section of the template.
%This section summarizes general information, such as the software name, purpose,
%platform, programming language, publications about the software, the first
%release and the most recent change date, website, source code repository of the
%product, number of developers, etc.  We only measure open source software that
%is ready for public use by other researchers (following the definition of the
%public category from \cite{GewaltigAndCannon2012}). Information in the summary
%section sets the context for the project, but it does not directly affect the
%grading scores.

%For measuring each quality, we ask several questions and the typical answers are
%among the collection of ``yes'', ``no'', ``n/a'', ``unclear'', a number, a
%string, a date, a set of strings, etc. 
The grader, after answering questions for each quality assigned an overall
score (between 1 and 10) based on the answers.  Several of the qualities use
the word ``surface'' to highlight that these particular qualities are a shallow
measure.  For example, we do not do user studies for measuring usability.
Instead, we are look signs that the developers considered usability.
We use two freeware tools to collect repository related data:
\href{https://github.com/tomgi/git_stats}{GitStats} and
\href{https://github.com/boyter/scc}{Sloc Cloc and Code (scc)}.  Further
details on quality measurement are provided in~% Smith et al.\ (2024)
\cite{SmithEtAl2024_MI_SOP}.

@@

Developed by Saaty in the 1970s, AHP is widely used to 
analyze multiple criteria decisions~\cite{VaidyaEtAl2006}. AHP organizes
multiple criteria in a hierarchical structure and uses pairwise comparisons
between alternatives to calculate relative ratios~\cite{Saaty1990}. AHP works
with sets of $n$ \textit{options} and $m$ \textit{criteria}.  In our project
$n=29$ and $m=9$ since there are 29 options (software products) and 9 criteria
(qualities). With AHP the sum of the grades (scores) for all products for a
given quality will be 1.0.  We rank the software for each of the qualities, and
then we combine the quality rankings into an overall ranking based on the
relative priorities between qualities.

@@

We initially identified 48 candidate software projects from
the literature \cite{Bjorn2017, Bruhschwein2019, Haak2015}, on-line articles
\cite{Emms2019, Hasan2020, Mu2019}, and forum discussions \cite{Samala2014}.
Then we filtered as follows:

@@

\item Removed the packages with no source code available, such as
\textit{MicroDicom}, \textit{Aliza}, and \textit{jivex}.

\item Focused on MI software that provides visualization functions.
We removed seven packages
that were toolkits or libraries, such as \textit{VTK}, \textit{ITK}, and
\textit{dcm4che}, and another three that were for PACS.

\item Removed \textit{Open Dicom Viewer} as it has not received any
updates since 2011.

@@

The Domain Expert provided a list 12 software packages.
We found 6 packages were on both lists: \textit{3D
Slicer}, \textit{Horos}, \textit{ImageJ}, \textit{Fiji}, \textit{MRIcron} 
(we use its descendant \textit{MRIcroGL}) and \textit{Mango} (we
use the web version \textit{Papaya}).  The remaining six packages were on our
out-of-scope list. The Domain Expert agreed with our final choice of 29
packages -- see Table~\ref{tab_final_list}. This contains summary data
collected in the year 2020.

The projects are sorted in descending order of lines of code.
We found the initial release dates (Rlsd) for most
projects and marked the two unknown dates with ``?''. The date of the last
update is the date of the latest update, at the time of measurement. We found
funding information (Fnd) for only eight projects.  For the Number Of
Contributors (NOC) we considered anyone who made at least one accepted commit as
a contributor. The NOC is not usually the same as the number of long-term
project members, since many projects received change requests and code from the
community.  With respect to the OS, 25 packages work on all three OSs: Windows
(W), macOS (M), and Linux (L). Although the usual approach to cross-platform
compatibility was to work natively on multiple OSes, five projects achieved
platform-independence via web applications. The full measurement data for all
packages is available on \href{https://data.mendeley.com/datasets/k3pcdvdzj2/1}
{Mendeley Data}.

@@

The programming languages used in order of decreasing popularity are \CC,
JavaScript, Java, C, Python, Pascal, Matlab.  The most popular language is \CC,
for 11 of 29 projects; Pascal and Matlab were each used for a single project.

@@

Figure \ref{fg_installability_scores} lists the installability scores.  We found
installation instructions for 16 projects, but two did not need them
(\textit{BioImage Suite Web} and \textit{Slice:Drop})
as they are web applications. 10 of the projects required extra
dependencies: Five depend on a specific browser; \textit{dwv}, \textit{OHIF
Viewer}, and \textit{GATE} needs extra libraries to build; \textit{ImageJ} and
\textit{Fiji} need an unzip tool; \textit{MatrixUser} needs Matlab;
\textit{DICOM Viewer} needs a Nextcloud platform.

@@

The scores are based on the ease of following the installation instructions,
and automated installation and uninstallation process. There were no issues
for all but the bottom four, just various degrees of ease and automation.

\textit{GATE}, \textit{dwv}, and \textit{DICOM Viewer} showed severe
installation problems. We were not able to install them, even after a reasonable
amount of time (2 hours).  For \textit{dwv} and \textit{GATE} we failed to build
from the source code, but we were able to proceed with measuring other qualities
using a deployed on-line version for \textit{dwv}, and a VM version for
\textit{GATE}. For \textit{DICOM Viewer} we could not install the NextCloud
dependency, and thus could not measure reliability nor robustness for it.

\textit{MatrixUser} depends on Matlab, whose installation is not easy.
For users who already have Matlab, this score should be higher.

@@

The packages with higher scores for correctness and verifiability (see
Figure~\ref{fg_correctness_verifiability_scores}) used a wider array of
techniques to improve correctness, and had better documentation to witness
this. For instance, we looked for evidence of unit testing, and only evidence
for about half of the projects. We identified five projects using CI/CD
tools: \textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji}, \textit{dwv}, and
\textit{OHIF Viewer}.

@@

Even for some projects with well-organized documentation, requirements
specifications and theory manuals were still missing.
The only requirements-related document we found was a road
map of \textit{3D Slicer}, which contained design requirements for upcoming
changes.

@@

Figure~\ref{fg_reliability_scores} shows the our results.  We were able to
follow the steps in the tutorials that existed (seven packages had them.)
However
\textit{GATE} could not open macro files and became unresponsive several times,
without any descriptive error message. We found that \textit{Drishti} crashed
when loading damaged image files, without showing any descriptive error message.

@@

Figure \ref{fg_robustness_scores} presents the scores for surface robustness.
The packages with higher scores elegantly handled unexpected/unanticipated
inputs, typically showing a clear error message. We may have underscored
\textit{OHIF Viewer}, since we needed further customization to load
data.
